---
title: "Simple logistic regression model"
author: "Tim D. Smith"
date: "July 11, 2015"
output: html_document
---

# Exploring the Titanic data set

Start by loading some libraries and loading the training data:
```{r message=FALSE}
library(dplyr)
library(ggplot2)
library(GGally)
train = read.csv("train.csv")
```

Let's plot the structured columns against each other and see what we learn:
```{r warning=FALSE,message=FALSE,fig.width=10,fig.height=10}
skipcols = c("Cabin", "Ticket", "Name", "PassengerId")
train$CabinLetter = as.factor(gsub("^([A-Z]).*", "\\1", train$Cabin))
train$Survived = factor(train$Survived, levels=c(0,1), labels=c("No", "Yes"))
train$Pclass = as.factor(train$Pclass)
train_to_plot = train[,!names(train) %in% skipcols]
print(ggpairs(train_to_plot) + theme_bw())
```

There's a lot going on here but we can extract some observations:

* Most passengers did not survive.
* The absolute number of survivors from each class is about the same, but there are many more nonsurvivors from passenger class 3.
* Women survived at a much higher rate.
* Age is not strongly linked to survival.
* Siblings & spouses are not linked to survival.
* More survivors had at least one parent or child than non-survivors, so there's some protective effect.
* More survivors paid a higher fare than not.
* Passengers embarking at Southhampton were less likely to survive.

Let's look at cabin letter more closely:
```{r}
fraction_yes = function(x) { sum(x == "Yes") / length(x) }
g = train %>%
  group_by(CabinLetter) %>%
  summarize(percent_survival=fraction_yes(Survived)) %>%
  ggplot(aes(CabinLetter, percent_survival)) +
    geom_bar(stat = "identity") +
    theme_bw()
print(g)
```

Survival was poor for poorly documented passengers and passengers in A and T cabins (though there was only one passenger in "T", which is probably an error in my heuristic for extracting deck, so let's not consider it).

# Logistic regression model

Let's build a simple logistic regression model. We will include the effects of:
* sex
* undocumented cabin and cabin A
* passenger class
* Southhampton embarkation

```{r}
prepare_for_fit = function(df) {
  train_logit = df
  train_logit$CabinLetter = as.factor(gsub("^([A-Z]).*", "\\1", train_logit$Cabin))
  train_logit$Pclass = as.factor(train_logit$Pclass)
  train_logit$BadCabin = train_logit$CabinLetter == "" | train_logit$CabinLetter == "A"
  train_logit$BadEmbarkation = train_logit$Embarked == "S"
  train_logit$ThirdClass = train_logit$Pclass == "3"
  train_logit
}

train_logit = prepare_for_fit(train)
train.model = glm(Survived ~ Sex + BadCabin + BadEmbarkation, data=train_logit, family="binomial")
print(summary(train.model))
```

We can ask naïvely how many of our passengers we classified correctly. (We need to partition our data-set into train and cross-validation segments to do this correctly...)

```{r}
sum((predict(train.model) > 0) == (train_logit$Survived == "Yes")) / length(train_logit$Survived)
```

Not great, but not bad, either.

# Apply the logistic regression model to our test data set and save a CSV

```{r}
test = read.csv("test.csv")
test_logit = prepare_for_fit(test)
test$Survived = predict(train.model, test_logit) > 0
output = data.frame(PassengerId=test$PassengerId, Survived=as.numeric(test$Survived))
write.csv(output, "logistic_model.csv", row.names=FALSE)
```

# Result

This scores 0.76077 which, as we expect, is a little worse than it performs on the training set. This is a significant improvement over the trivial solution where we predict no survivors, which scores 0.62679.

# Adding fare data

We expect fare data to have a nonlinear relationship with survival probability, so let's see if we can find a threshold score above which survival is most likely. We can do this using a ROC curve for fare vs. survival.

```{r message=FALSE}
library(pROC)
roc1 = roc(Survived ~ Fare, train,
           plot=TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, grid=TRUE,
           print.auc=TRUE, show.thres=TRUE, print.thres="best")
```

So let's add that to our model and try again.

```{r}
train_logit$Expensive = train_logit$Fare > 10.825
train.model = glm(Survived ~ Sex + BadCabin + BadEmbarkation + Expensive, data=train_logit, family="binomial")
print(summary(train.model))
print(sum((predict(train.model) > 0) == (train_logit$Survived == "Yes")) / length(train_logit$Survived))
```

It looks like this gives us only marginal improvement over our prior model, but let's run it through the leaderboard anyway.

```{r}
test_logit$Fare[is.na(test_logit$Fare)] = median(test_logit$Fare, na.rm=TRUE)
test_logit$Expensive = test_logit$Fare > 10.825
test$Survived = predict(train.model, test_logit) > 0
output = data.frame(PassengerId=test$PassengerId, Survived=as.numeric(test$Survived))
write.csv(output, "logistic_model_fare.csv", row.names=FALSE)
```

This analysis makes us a little more pessimistic about the fate of two of our passengers:
```
tim@rocketman:titanic (master *)$ diff -U0 logistic_model.csv logistic_model_fare.csv 
--- logistic_model.csv	2015-07-11 12:31:53.000000000 -0700
+++ logistic_model_fare.csv	2015-07-11 13:03:57.000000000 -0700
@@ -290 +290 @@
-1180,1
+1180,0
@@ -323 +323 @@
-1213,1
+1213,0
```

and actually scores worse than our simpler model, with a score of 0.75598. I think this is likely because the effect of fare was already encompassed by the effect of cabin and class.